---
title: "Assessing MVP Votes by WAR"
author: "Dean Wise"
date: "8/7/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
war_file <- "~/Desktop/R Scripts/Personal/Underrated/war_daily_bat.csv"
```

WAR (Wins Above Replacement) is often looked at as a standardized measure of a baseball players contributions: The higher the WAR the higher the value they represent for the team. While WAR has its limitations, we are going to look at how it compares to a players MVP votes during that season. Mores pecifically, we are going to look at WAR compared to the share of MVP votes (total votes earned / total possible votes earned) 


First we are going to load relevant packages and then load in our WAR data from Baseball Reference (http://www.baseball-reference.com/data/war_daily_bat.txt)

```{r load_war, message=FALSE, warning=FALSE}
needs(tidyverse, Lahman, DT, lubridate, plotly)
#Note I have already saved down the War file onto my local machine
war <- read_csv(war_file)
#Lets see a random sample of the data table
datatable(sample_n(war, 50))
```
***

Let's now clean up the file and take the relevant columns we need:
```{r clean_war}
clean_war <- war %>% 
  select(player_ID, name_common, year_ID, team_ID, WAR)
#Take a look at what we now have
datatable(sample_n(clean_war, 5))
```

***

Next, Lets try to gather a list of all MVP winners from the AwardsPlayers table in the Lahman database. We are going to try and get the PlayerID, the Year, and then add in a column where we indicate they won the award that year
```{r mvp_winners}
data(AwardsPlayers)
won <- AwardsPlayers %>% 
  filter(awardID == "Most Valuable Player") %>% 
  select(playerID, yearID) %>% 
  mutate(Won = "Won")
datatable(won)
```

***

Let's try to remove pitchers since their WAR does not really match those of a non-pitcher:
```{r pitcher_list}
data(Pitching)
pitchers <- Pitching %>% 
  group_by(playerID) %>% 
  summarise(games = sum(G)) %>% 
  filter(games>=5)
pitchers <- pitchers$playerID
```


Now, lets build a table that brings in the MVP votes for each player and joins that with the WAR for that year, along with if they won an MVP award or not

```{r build_main}
data(AwardsSharePlayers)
mvp <- AwardsSharePlayers %>% 
  filter(awardID == "MVP") %>% 
  select(playerID, yearID, lgID, pointsWon, pointsMax, votesFirst) %>% 
  left_join(clean_war, by = c("playerID" = "player_ID", "yearID" = "year_ID")) %>% 
  left_join(won, by = c("playerID" = "playerID", "yearID" = "yearID")) %>% 
  mutate(Won = if_else(is.na(Won) == TRUE, "Did not Win MVP", "Won MVP"), 
         Player = paste(name_common, " - ", as.character(yearID), " (", team_ID, ")", sep = ""), 
         Share_of_Vote = pointsWon/pointsMax) %>% 
  filter(!(playerID %in% pitchers)) %>% 
  select(Player, yearID, lgID, WAR, Share_of_Vote, Won) %>% 
  filter(is.na(WAR) == FALSE & yearID >=1980) %>% 
  mutate(WAR = as.double(WAR), 
         Share_of_Vote = as.double(Share_of_Vote))

datatable(mvp)
```

```{r, message=FALSE, warning=FALSE, include=FALSE}
war <- NULL
clean_war <- NULL
won <- NULL
```

***

Lets plot these players:

```{r plotly, message=FALSE, warning=FALSE}
p <- plot_ly(mvp, x = ~WAR, y = ~Share_of_Vote, type = "scatter", text = ~Player, color = ~Won, width = 1100, height = 500)
p
```

***

Findings
---

Some things that immidiately stand out:

* Barry Bonds was an insane baseball specimen in the early 2000's.

* Mike Trout in 2012 was the most deserving MVP to not earn one with a WAR of 10.75.

* The Big Hurt (Frank Thomas) earned himself 100% of the votes in '93 yet had a surprisingly low WAR of just over 6. 

* Cal Ripken had a share of Vote of 0.2% in '84 despite having a WAR of 10

